\textbf{Efficient Implementation Strategies for Block Ciphers on ARMv8}\\ {\footnotesize Bachelorarbeit}

### 2023-03-13 Paper/paper.tex
encryption working

We choose $n=m=1$, so for each encryption, we divide eight messages into four
registers.

@@

Let $a,b,c,d,e,f,g,h$ be eight messages of length $64$ with subscripts denoting
individual bits. We first put these messages into four SIMD registers
$V_0,V_1,V_2,V_3$:

\begin{align*}
    V_0&=b||a \\
    V_1&=d||c \\
    V_2&=f||e \\
    V_3&=h||g
\end{align*}

We then use the \texttt{SWAPMOVE} technique to bring the data into bitslice
format. This operation operates on two registers $A,B$ using mask $M$ and shift
value $N$. It swaps bits in $A$ masked by $M$ with bits masked by $M\ll N$ in
B using only three XOR-, one AND- and two shift operations.

@@

The following operations put all of the first bits of each S-Box into $V_0$,
each second bit into $V_1$ and so forth and regroups these bits into bytes.

\begin{align*}
    &\text{SWAPMOVE}(V_0,V_1,0x5555\dots 55,1) &\text{SWAPMOVE}(V_0,V_0,0xf0f0\dots 0f,4) \\
    &\text{SWAPMOVE}(V_2,V_3,0x5555\dots 55,1) &\text{SWAPMOVE}(V_1,V_1,0xf0f0\dots 0f,4) \\
    &\text{SWAPMOVE}(V_0,V_2,0x3333\dots 33,2) &\text{SWAPMOVE}(V_2,V_2,0xf0f0\dots 0f,4) \\
    &\text{SWAPMOVE}(V_1,V_3,0x3333\dots 33,2) &\text{SWAPMOVE}(V_3,V_3,0xf0f0\dots 0f,4)
\end{align*}

With numbers denoting byte groups (i.e. $3$ corresponding to
$h_3g_3f_3e_3d_3c_3b_3a_3$), our data now has the following
permutation-friendly format:

\begin{alignat*}{7}
    V_0&=&&124&&||\dots&&||12&&||8 &&||4&&||0 \\
    V_1&=&&125&&||\dots&&||13&&||9 &&||5&&||1 \\
    V_2&=&&126&&||\dots&&||14&&||10&&||6&&||2 \\
    V_3&=&&127&&||\dots&&||15&&||11&&||7&&||3 \\
\end{alignat*}

@@

One thing to take note of is the original permutation values only showing where
a given byte should land, not which byte belongs to a certain position - i.e.
for $P_0$, byte 1 should land in position 12, but the byte belonging to
position 1 is byte 4. Because \texttt{vperm} works in the latter way, we have
to do some trivial rearrangements. Putting these values into registers
$V_5,V_6,V_7,V_8$ now allows us to compute the permutation layer in only four
instructions:

\begin{align*}
    &\texttt{TBL V0, V0, V5} &\texttt{TBL V1, V1, V6} \\
    &\texttt{TBL V2, V2, V7} &\texttt{TBL V3, V3, V8}
\end{align*}

>>>>

$n=m=1$ would be a valid choice which yields eight messages divided into four
registers. We choose $n=m=2$ so we can directly utilize the algorithm for bit
packing presented by the original GIFT authors, although it is simple to adapt
this algorithm to only four registers and eight messages by dividing each
128-bit NEON register into two separate 64-bit registers and adjusting the
\texttt{SWAPMOVE} shift and mask values.

@@

Let $a,b,\dots,p$ be sixteen messages of length $64$ with subscripts denoting
individual bits. We first put these messages into eight SIMD registers
$V_0,V_1,\dots,V_7$:

\begin{alignat*}{3}
    V_0&=b||a\qquad V_4&&=j||&&i \\
    V_1&=d||c\qquad V_5&&=l||&&k \\
    V_2&=f||e\qquad V_6&&=n||&&m \\
    V_3&=h||g\qquad V_7&&=p||&&o
\end{alignat*}

We then use the \texttt{SWAPMOVE} technique to bring the data into bitslice
format. This operation operates on two registers $A,B$ using mask $M$ and shift
value $N$. It swaps bits in $A$ masked by $(M\ll N)$ with bits in $B$ masked by
$M$ in using only three XOR-, one AND- and two shift operations.

@@

The following operations group all $i$th bits of the messages $a,c,\dots,o$
into bytes and puts these into the lower half of the registers $V_{i\bmod 8}$.
The same is done for messages $b,d,\dots,p$, only differing in that the bytes
are put into the upper half of the registers.

\begin{align*}
    &\text{SWAPMOVE}(V_0,V_1,0x5555\dots 55,1) &\text{SWAPMOVE}(V_4,V_5,0x5555\dots 55,1) \\
    &\text{SWAPMOVE}(V_2,V_3,0x5555\dots 55,1) &\text{SWAPMOVE}(V_6,V_7,0x5555\dots 55,1) \\
    &\text{SWAPMOVE}(V_0,V_2,0x3333\dots 33,2) &\text{SWAPMOVE}(V_4,V_6,0x3333\dots 33,2) \\
    &\text{SWAPMOVE}(V_1,V_3,0x3333\dots 33,2) &\text{SWAPMOVE}(V_5,V_7,0x3333\dots 33,2) \\
    &\text{SWAPMOVE}(V_0,V_4,0x0f0f\dots 0f,4) &\text{SWAPMOVE}(V_1,V_5,0x0f0f\dots 0f,4) \\
    &\text{SWAPMOVE}(V_2,V_6,0x0f0f\dots 0f,4) &\text{SWAPMOVE}(V_3,V_7,0x0f0f\dots 0f,4) \\
\end{align*}

With $Ax=o_xm_xk_xj_xg_xe_xc_xa_x$ and $Bx=p_xn_xl_xi_xh_xf_xd_xb_x$ denoting
byte groups, our data now has the following permutation-friendly format:

\[
    \begin{array}{c|llllllll}
        n & 7 & 6 & 5 & 4 & 3 & 2 & 1 & 0 \\
        \hline
        V_0 & A56 & A48 & A40 & A32 & A24 & A16 & A8 & A0 \\
        V_1 & A57 & A49 & A41 & A33 & A25 & A17 & A9 & A1 \\
        V_2 & A58 & A50 & A42 & A34 & A26 & A18 & A10 & A2 \\
        V_3 & A59 & A51 & A43 & A35 & A27 & A19 & A11 & A3 \\
        V_4 & A60 & A52 & A44 & A36 & A28 & A20 & A12 & A4 \\
        V_5 & A61 & A53 & A45 & A37 & A29 & A21 & A13 & A5 \\
        V_6 & A62 & A54 & A46 & A38 & A30 & A22 & A14 & A6 \\
        V_7 & A63 & A55 & A47 & A39 & A31 & A23 & A15 & A7
    \end{array}
\]

\[
    \begin{array}{c|llllllll}
        n & 15 & 14 & 13 & 12 & 11 & 10 & 9 & 8 \\
        \hline
        V_0 & B56 & B48 & B40 & B32 & B24 & B16 & B8 & B0 \\
        V_1 & B57 & B49 & B41 & B33 & B25 & B17 & B9 & B1 \\
        V_2 & B58 & B50 & B42 & B34 & B26 & B18 & B10 & B2 \\
        V_3 & B59 & B51 & B43 & B35 & B27 & B19 & B11 & B3 \\
        V_4 & B60 & B52 & B44 & B36 & B28 & B20 & B12 & B4 \\
        V_5 & B61 & B53 & B45 & B37 & B29 & B21 & B13 & B5 \\
        V_6 & B62 & B54 & B46 & B38 & B30 & B22 & B14 & B6 \\
        V_7 & B63 & B55 & B47 & B39 & B31 & B23 & B15 & B7
    \end{array}
\]

@@

One thing to take note of is the original permutation values only showing where
a given byte should land, not which byte belongs to a certain position - i.e.
for $P_0$, byte 1 should land in position 12, but the byte belonging to
position 1 is byte 4. Because \texttt{vperm} works in the latter way, we have
to do some trivial rearrangements. In addition, because data for each slice is
split between two registers $(V_0,V_4),(V_1,V_5),\dots$, the \texttt{vperm}
operation needs to use these two registers as a data source. Using two data
sources is supported by the \texttt{TBL} instruction by means of an additional
parameter.

Assuming the correct permutation values are put into registers
$V_9,V_{10},\dots,V_{16}$, this now allows us to compute the permutation layer for
all 16 blocks in only eight permutation instructions plus four additional vector
copy instructions used for saving the original values of $V_0,V_1,V_2,V_3$
before the permutation is applied.

\begin{alignat*}{2}
    &\texttt{MOV V17, V0}\qquad &&\texttt{MOV V18, V1} \\
    &\texttt{MOV V19, V2}\qquad &&\texttt{MOV V20, V3} \\
    &\texttt{TBL V0, \{V0, V4\}, V9}\qquad &&\texttt{TBL V1, \{V1, V5\}, V10} \\
    &\texttt{TBL V2, \{V2, V6\}, V9}\qquad &&\texttt{TBL V3, \{V3, V7\}, V10} \\
    &\texttt{TBL V4, \{V17, V4\}, V9}\qquad &&\texttt{TBL V5, \{V18, V5\}, V10} \\
    &\texttt{TBL V6, \{V19, V6\}, V9}\qquad &&\texttt{TBL V7, \{V20, V7\}, V10}
\end{alignat*}
